Gravitational field calculation in spherical coordinates using variable densities in depth

### 2019-03-09 manuscript/manuscript.tex
Add sinusoidal disclaimer in discussion
Move disclaimer from discussion to conclusions
Change Appendix A: shell solutions from Newton

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting two dimensional adaptive discretization algorithm that
reduces the errors due to the distance between the tesseroid and the computation point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear,
exponential, and sinusoidal density functions.
The heavily oscillating density functions are intended only to test the algorithm to its
limits and not to emulate a real world case.
These comparisons allowed us to obtain optimal values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential and
its gradient are 1 and 2.5, respectively.
The density-based discretization algorithm produces no discretizations in the linear
density case, but a delta ratio of 0.1 is needed for the exponential and the
sinusoidal density functions.
These values can be extrapolated to cover most common use cases.
However, the distance-size and delta ratios can be configured by the user to increase
the accuracy of the results at the expense of computational speed.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000~m, using an exponential density function.
\end{summary}

@@

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006, Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2018} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
We derive an expression for the gravitational potential of a spherical shell with
an arbitrary density function in depth (Eq.~\ref{eq:shell-pot}),
from which we obtained
expressions for linear, exponential, and sinusoidal density functions
(see Appendix~\ref{sec:shell}).

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{figures/spherical-shell.pdf}
\caption{
    Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
    The computation point $Q$ is located on the $z$ axis at a distance $r$ from
    the origin of the coordinates system.
    For our purposes we will assume that $Q$ is outside of the shell,
    i.e.~$r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

Consider a spherical shell with inner radius $R_1$ and outer radius $R_2$,
whose density is a function $\rho(r')$ of the radial coordinate
(Fig.~\ref{fig:spherical-shell}).
The gravitational potential generated by the shell on an arbitrary external
point $Q$ can be written as follows:

\begin{equation}
    V_\text{sh}(\phi, \lambda, r) = G
    \int\limits_0^{2\pi}
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r')}{\ell} {r'}^2 \cos\phi' \,
    dr' d\phi' d\lambda',
\end{equation}

\noindent where $\ell$ is defined in equation~\ref{eq:ell}.

For $Q$ located along the $z$ axis (i.e., $\phi=90^\circ$) at a distance $r$ from the
origin, equation~\ref{eq:ell} simplifies to:

\begin{equation}
    \ell = \sqrt{r'^2 + r^2 - 2 r r' \sin\phi'}.
\end{equation}

\noindent
Because of the rotational symmetry along the $z$ axis, the integration in $\lambda'$ is
straightforward:

\begin{equation}
    V_\text{sh}(r) = 2\pi G
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r') {r'}^2 \cos\phi'}{\sqrt{r'^2 + r^2 - 2 r r' \sin\phi'}}
    \, dr' d\phi',
\end{equation}

\noindent
while the integration in $\phi'$ can be performed independently of the density function.
Making use of SymPy \citep{sympy2017}, a Python library for symbolic mathematics, we
obtained the following expression for the potential:

\iftwocol{
\begin{equation}
    \begin{split}
        V_\text{sh}(r) = 2\pi G
        \int\limits_{R_1}^{R_2}
        \Big[ & \sqrt{r^2 + r'^2 + 2rr'} - \\
        & \sqrt{r^2 + r'^2 - 2rr'}
        \Big] \frac{r'\rho(r')}{r} \, dr'.
    \end{split}
\label{eq:shell-pot-sqrts}
\end{equation}
}{
\begin{equation}
    V_\text{sh}(r) = 2\pi G
    \int\limits_{R_1}^{R_2}
    \Big[ \sqrt{r^2 + r'^2 + 2rr'}  -
    \sqrt{r^2 + r'^2 - 2rr'}
    \Big] \frac{r'\rho(r')}{r} \, dr'.
\label{eq:shell-pot-sqrts}
\end{equation}
}

Because the computation point $Q$ is outside of the shell, $r > r'$ and the square roots
in equation~\ref{eq:shell-pot-sqrts} simplify to

\begin{equation}
    \sqrt{r^2 + r'^2 + 2rr'} = |r + r'| = r + r',
\end{equation}

\begin{equation}
    \sqrt{r^2 + r'^2 - 2rr'} = |r - r'| = r - r',
\end{equation}

\noindent which leads to the following expression for the potential:

\begin{equation}
    V_\text{sh}(r) = \frac{4\pi G}{r}
    \int\limits_{R_1}^{R_2} {r'}^2 \rho(r') \, dr'.
\label{eq:shell-pot}
\end{equation}

@@

From Eq.~\ref{eq:shell-pot} we can obtain expressions of the gravitational potential for
different density functions.

>>>>

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting two dimensional adaptive discretization algorithm that
reduces the errors due to the distance between the tesseroid and the computation point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear,
exponential, and sinusoidal density functions.
These comparisons allowed us to obtain optimal values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential and
its gradient are 1 and 2.5, respectively.
The density-based discretization algorithm produces no discretizations in the linear
density case, but a delta ratio of 0.1 is needed for the exponential and the
sinusoidal density functions.
These values can be extrapolated to cover most common use cases.
However, the distance-size and delta ratios can be configured by the user to increase
the accuracy of the results at the expense of computational speed.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000~m, using an exponential density function.
\end{summary}

@@

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006, Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2018} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
Applying the Newton's Shell Theorem \citep{Chandrasekhar1995, Binney2008},
we derive expressions for the gravitational potential of a spherical shell with
linear, exponential, and sinusoidal density functions
(see Appendix~\ref{sec:shell}).

@@

Consider a spherical shell with inner radius $R_1$ and outer radius $R_2$,
whose density is a function $\rho(r')$ of the radial coordinate.
We want to get an analytical expression for the gravitational fields the shell generates
on any external point located at a distance $r$ from the centre of the shell
($r > R_2$).

According to Newton's Shell Theorem (Theorem XXXI)
\citep{Chandrasekhar1995, Binney2008}, the gravitational potential generated by the
shell on any external point is the same as it would be if its mass were concentrated on
a point located at its centre:

\begin{equation}
    V_\text{sh}(\phi, \lambda, r) = \frac{GM}{r},
\end{equation}

\noindent where $M$ is the total mass of the shell, which can be easily computed as
follows:

\begin{equation}
    M =
    \iiint\limits_{\Omega} \rho(r') dV =
    4\pi \int\limits_{R_1}^{R_2} \rho(r') {r'}^2 dr',
\end{equation}

\noindent where $\Omega$ symbolizes the volume of the shell.

Combining the two previous equations, we get the following expression for the potential:

\begin{equation}
    V_\text{sh}(r) = \frac{4\pi G}{r}
    \int\limits_{R_1}^{R_2} {r'}^2 \rho(r') \, dr',
\label{eq:shell-pot}
\end{equation}

\noindent which is in agreement with the one obtained by \citet[p.62]{Binney2008}.

@@

From Eq.~\ref{eq:shell-pot} we can obtain expressions of the gravitational potential for
different density functions.
The integration of the following density functions have been carried out by SymPy
\citep{sympy2017}, a Python library for symbolic mathematics.
